Depolarization of Interacting NV ensembles : \\ The Role of Local Electric Fields and Double Flip processes

### 2022-06-22 Mes papiers/Low field CR/Low_field_CR.tex
tutu

The electronic spin properties of the negatively charged nitrogen-vacancy (NV$^-$) center in diamond has given rise to a wealth of applications in nanoscale sensing and quantum information science in part thanks to the possibility to optically polarize and read-out its spin state at ambient conditions. Many recent work indeed focus on ensembles of spins from NV centers with the prospect to boost the magnetic field sensing capabilities and for studying many body effects (resultats de chez BOSCH, LUKIN , FEFOR, ALEX, BIFONE)

When the spin ensemble becomes dense, inhomogeneities in the relaxation rates of NVs can give rise to spin depolarisation with a very rich many body dynamics associated with disorder. This mechanisms however limits the efficiency of magnetometers (LUKIN PRX).
%It was proposed to exploit these CR to perform sensitive microwave free magnetometry (REF).  
Recent magnetometry proposals have been put forward at zero magnetic field. There, the depolarisation effect was shown to be much stronger, but the detailed studies of relaxation mechanisms have not been identified to the best of our knowledge. Some of the possible processes that lead to cross-relaxation in strongly coupled dipolar systems are depicted in Fig.~1. The goal of this paper is to identify regimes where the spin flip-flop within different classes of NV centers, double-flip up and down as well as the mixing induced by local electric field play a role.

The electronic spin of NV center is a spin-1 system in the ground state (see Fig. 1). It can be optically polarized in the $\ket{m_s=0}$ state. The photoluminescence of this state is also is larger than the $\ket{m_s=\pm 1}$ states enabling spin-read out at ambient temperature. The $\ket{m_s=\pm 1}$ spin states are separated from the $\ket{m_s=0}$ state by  $D = (2\pi) 2.87$ GHz so that when a resonant microwave or static transverse magnetic field is applied, the photoluminescence is reduced \citep{epstein2005anisotropic,lai2009influence}. 
Fig.  \ref{T1}-a) and b) show the change in photoluminescence (PL) with respect to an external magnetic field for two samples with a low and high concentration of NV$^-$ centers (see SI). Both samples show a decrease in PL as the magnetic field amplitude increases.  There is however a stark difference in the low magnetic field region where only high-density samples shows a drop in PL (CITATION). This effect can be observed on all our samples (see SI) whenever the NV concentration lies in the ppm range.

The drop of the PL at low magnetic field is generally associated with a decrease of the NV's spin lifetime $T_1$ (CITE JARMOLA), which results in a decrease of the population in the bright state $0$.
%Indeed, since the $\ket{m_s=0}$ spin state is brighter than the $\ket{m_s=\pm 1}$ states, and since the optical pumping in the $\ket{m_s=0}$ state is in competition with the relaxation in a thermal distribution of the states, decreasing the spins lifetime results in a decrease of the NV PL \citep{finco2021imaging}.
%To understand the behavior of the PL in zero-field, we therefore need to study the dynamics of the spin in the low magnetic field region.
The depolarization dynamics of the spin of single or dilute NV centers at room temperature is dominated by two-phonon Raman processes \citep{redman1991spin,jarmola_temperature-_2012,norambuena2018spin}, which depend on the crystal lattice temperature. It has been observed that dense NV centers ensemble have an additional spin decay channel \citep{jarmola_temperature-_2012,jarmola_longitudinal_2015,mrozek_longitudinal_2015, choi_depolarization_2017, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, pellet2021magnetic, mrozek2021characterization}, which depends greatly on the magnetic field amplitude. This effect has been attributed to cross-relaxation between the NV centers through dipole-dipole coupling \citep{mrozek_longitudinal_2015, choi_depolarization_2017}.  Inhomogeneity of the spin lifetimes is further needed in order to explain the depolarization of the spin ensemble. We will denote $T_1^{\rm ph}$ the characteristic timescale associated with the phonon relaxation process and $T_1^{\rm dd}$ the density dependent timescale associated with the dipole-dipole relaxation process.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/shema_summary.pdf}
\caption{Schematics showing a diamond with interacting spins as well as the three processes that can account for dipolar relaxation. }
\label{T1}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig_T1.pdf}
\caption{a) Photoluminescence measurement from NV centers ensemble as a function of an external magnetic field applied in an arbitrary direction (a) for a CVD sample containing $\approx4$ ppb of NV$^-$ centers, (b) for an HPHT sample containing $\approx3$ ppm of NV$^-$ centers. 
(c) Sequence used to measure the spin lifetime. (d) Spin relaxation $S_1-S_2$ measured for the dense sample at zero and non-zero magnetic fields. The fitting procedure (orange plain line) is detailed in the main text}
\label{T1}
\end{figure}

@@

Fig. \ref{T1}-c) shows the sequence employed for measuring $T_1^{\rm dd}$ in the dense sample. 
%The spin lifetime protocol used here is described in Fig. \ref{T1} and is based on previous similar experiments 
It consists in a pump-probe measurement where the spins are first polarized in the $\ket{0}$ state by a green laser, and read-out optically after a variable dark time $\tau$. In highly doped samples, this sequence often results in artifacts, mostly due to charge state transfer in the dark \citep{giri_coupled_2018, giri_selective_2019}. It is therefore convenient to repeat the sequence with an additional $\pi$ pulse right before the spin read-out to prepare the remaining $\ket{0}$ polarization into a darker $\ket{+1}$ or $\ket{-1}$ state. By subtracting the result of the two sequences, we select only the spin-dependent part of the signal, with the added benefit of being able to select a specific class of NV centers \citep{jarmola_temperature-_2012, mrozek_longitudinal_2015, choi_depolarization_2017}.

Fig. \ref{T1}-b) shows the subtracted signals when $\bm B=0$ and $\bm B \neq 0$.
The ensemble spin decay curves are not exponential, which was found to be due to inhomogeneities between the spin lifetimes of dipolar coupled NV$^-$ centers \citep{choi_depolarization_2017}.
We will base our interpretations of the experimental results on the NV-fluctuator model developed in \citep{choi_depolarization_2017}. A conclusion of this model is that, for a homogeneous 3D distribution of fluctuators, the dipole-induced lifetime should be stretched-exponential with a stretch factor $\beta=1/2$. Such an ensemble lifetime is indeed observed when $T_1^{\rm dd} \ll T_1^{\rm ph}$.
For the samples used in this paper (detailed in SI) $T_1^{\rm dd} \sim T_1^{\rm ph}$, so both decay processes have to included in the analysis. Fitting the two curves by the product of an exponential and a stretched exponential, we find $T_1^{\rm ph}=3.2$ ms and $T_1^{\rm dd}= ??$ here. We also conducted further measurements that confirm the lifetime-limited character of the fluctuator employed in the model \citep{choi_depolarization_2017}, see SI. 
%These results show that our sample clearly manifests dipolar interactions amongst closely packed NV centers, but do not tell about the magnitude of the several possible mechanisms involved in the dipolar relaxation. 
%In spin-1 systems, there can indeed be several channels for decay at zero fields such as flip-flop, double flip,

@@

One of the ways to tune dipole-dipole related phenomena is by changing the number of resonant spins. Fig. \ref{100_VS_1x4} (b)-i) shows an optically detected magnetic resonance (ODMR) spectrum from the dense ensemble. The observed lines correspond to the projections of the magnetic fields onto the four "classes" of NV centers - that is physical orientation of the NV axis is the diamond crystal cell - together with the two possible spin transitions $\ket{0}\to\ket{-1}$ and $\ket{0}\to\ket{+1}$. The transitions frequencies of the four different classes can be controlled by changing the amplitude and orientation of the external magnetic field. Fig. \ref{100_VS_1x4} (c)-i) shows an example where all four classes are brought to resonance by placing the magnetic field along the [100] crystalline axis. 
%Other example of class resonances are given in the SI.
This configuration will be essential to decipher what dipolar relaxation mechanisms play a role at low field.

Fig. \ref{100_VS_1x4} (b)-ii) and (b)-iii) show the PL from an ensemble of NV centers as a function of the amplitude of a magnetic field aligned as in Fig. \ref{100_VS_1x4} (b)-i). 
We also extract the stretched exponential spin decay time $T_1^{\rm dd}$ by running the sequence shown in Fig. 2-c) at each of the XX magnetic field values. 
We observe that the 3 \% increase in PL under low magnetic field is indeed correlated with a drop in the spin decay rate from 1.6 kHz to 0.05 kHz, as the magnetic field increases. This result can be explained by the higher number of resonant spins in zero magnetic field, where all four classes are degenerate, as opposed when $\bm B \neq 0$. The half width of the dip ($\approx 10$G) is consistent with a fluctuator model taking into account the change in the spin resonance overlaps as the $B$ field is swept (see SI ?).
Note that the decrease in PL for $B>40\ \rm G$ are related to state mixing by the transverse magnetic field, so it is not correlated to a modification of the spin lifetime.

Importantly, we observed that the lifted degeneracy of the four NV classes is not the sole reason for the increase of the spin lifetime as the B field increases. Fig. \ref{100_VS_1x4} (c)-ii) and (c)-iii) show the change in the PL and in $1/T_1^{\rm dd}$ as a function of a magnetic field that is aligned along the [100] axis. For this particular orientation, the four classes are always resonant, regardless of the field amplitude. We can see that there still is an increase in the spin decay rate and a corresponding drop in the PL under low magnetic field values, although considerably reduced. Also, the rate of change in both the PL and $T_1^{\rm dd}$ is larger by more than a factor of 3 in the [100] configuration. Although the contrast is smaller, this effect can thus drastically impact zero-field magnetometry protocols based on CR.
The main aim of this paper is to understand the physics governing the contribution to the spin decay in this second scenario. Note that the slight drop in PL and the corresponding bump for $1/T_1^{\rm dd}$ at $B \sim 20$G is related to dipolar interaction with NV centers that have a $^{13}$C as a first neighbor \cite{pellet2021optical}.

We identified and isolated two additional mechanisms that could explain this effect : local electric fields and double-flip processes. 
The dipole-dipole interaction Hamiltonian between two spins ${\vec S}_1$ and ${\vec S}_2$ reads :
\begin{equation}
\mathcal{H}^{\rm dd}= -\frac{J_0}{r^3}\left(3\left({\vec S}_1 \cdot \vec u \right)\left({\vec S}_2 \cdot \vec u \right) - {\vec S}_1 \cdot {\vec S}_2  \right),
\end{equation}
Where $J_0= (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\vec r$ is the relative positions of the two spins and $\vec u = \frac{\vec r}{\norm{\vec r}}$. In situations such as the one described in Fig.XX, the only relevant terms of $\mathcal{H}^{\rm dd}$ in terms of population transfer are the flip-flop terms such as $\mel{0,+1}{\mathcal{H}^{\rm dd}}{+1,0}$. In zero magnetic field however, we have to take into account these two other mechanisms of the dipolar Hamiltonian in consideration.
%where both NV spins involved in the dipole-dipole coupling flip their spin in the same direction.
It was shown in  \cite{mittiga2018imaging} that local electric fields coming from charge traps such as $P_1^+$ centers are responsible for the optically detected spin resonance profile at zero magnetic field in large NV ensembles. Magnetic field noise comes in to second order in zero field. %We also do not take into account the hyper-fine structure of the NV center due to the large inhomogeneous broadening of the ESR transitions (see SI). 
We will then consider the following spin Hamiltonian for the electric field dependent NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_{\rm elec}= d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
where $E_{x,y}$ are the projections of local electric fields on the NV axes. 
Under local electric fields with orientations given by the angle $\phi_E={\rm tan}(E_x/E_y)$, the eigenstates of $H_{NV}$ are $\ket{0}$ and $\ket{\pm}=\frac{1}{\sqrt{2}}(\ket{+1}\pm e^{-i\phi_E}\ket{-1})$.
Computing the flip-flop terms $\ket{\pm,0} \bra{0,\pm} $ in the fluctuator model shows that $1/T_1^{dd}$ should be increased at zero field, even after averaging over all possible orientations $\phi_E$ (see SI).  Another mechanism at stake when considering dipolar interactions with spin-1 systems are double-flip processes. 
These processes correspond to terms like $\ket{+1,0} \bra{0,1}$ in the anisotropic dipolar interaction, giving rise to an exchange of two units of spin-angular momentum. These processes can become significant when the states $\pm 1$ are degenerate and are thus likely to explain part of the zero-field dip in Fig. c)-ii). 
%Theoretical model predicts a factor of $\approx $5 increase in $1/T_1^{dd}$ at zero field.

@@

Crucially, mixing remains even when the associated energy splitting of the $\ket{d}$ and $\ket{e}$ states is large : Fig. \ref{B_transverse}-(b) shows both the splitting between the energy levels of $\ket{d}$ and $\ket{e}$, and the matching factor $\abs{\bra{e}\ket{+}}^2$ characterizing the closeness of the states $\bra{e}$ and $\bra{+}$. We can see that at a field value of $\sim 130$ G, the splitting between the $\ket{d}$ and $\ket{e}$ reaches $\sim 50$ MHz, far exceeding the range of the dipole-dipole resonant coupling (see details in SI), but $\abs{\bra{e}\ket{+}}^2 > 0.98$ meaning that the spin Hamiltonian eigenstates are still roughly in the $\ket{0}, \ket{+}, \ket{-}$ basis. We can therefore decouple the effects of the double-flip processes  $\propto \ket{g,d}\bra{d,e}$, which are no longer resonant when $| \bm B | > 130$ G, from the effects of the mixing induced by the transverse electric - or in this case magnetic - field.

Fig. \ref{B_transverse}-(c) shows the measurement of the stretched lifetime $T_1^{\rm dd}$ for a single class of NV centers exposed to a transverse magnetic field between 25 (??) and 130 G (FOR A DIFFERENT SAMPLE... which ? we observe a $1/T_1^{\rm dd}$ of $30$ Hz under large longitudinal magnetic fields for this class ?). The detuning $\Delta \nu$ between the two states $\ket{d}$ and $\ket{e}$ was measured through ODMR. We attribute the decrease in the spin decay rate when the detuning increases to the loss of effectiveness of the double-flip processes. We can notice that the decay rate reaches a plateau for $\Delta \nu \geq 30\ \rm MHz$ with a value about twice as large as the decay rate for a longitudinal magnetic field. This plateau we attribute to the effect of the transverse magnetic field, which we assume to be similar to the effect of the electric field in the low magnetic field regime. We should note that the effect of the double-flip processes is about 4 times more important in the depolarization rate than the effect of the electric field, even though the states are not initially fully resonant.

@@

Fig. \ref{magneto} (a) shows the demodulated PL as a DC magnetic field is scanned in a random direction and an additional alternating magnetic field of a few Gauss at $\sim \ \rm kHz$ is added through the same electromagnet (see experimental details in SI) WHICH SAMPLE ? . We can see a sharp linear slope in low field $\abs{B} < 5\ \rm G$. Once calibrated, in this case with ODMR, the slope can provide a 1D magnetic field measurement, which could be extended to 3D with a set of 3 coils or 3 electromagnets, as in \cite{zheng_microwave-free_2020}. In order to assess the sensitivity of the measurement, we alternate a small DC field of $\approx 40\ \mu\rm{T}$ every few seconds and take a histogram of the measured fields, as shown in Fig. \ref{magneto} (b) and (c). The histogram is well fitted with gaussians of standard deviation $\sigma=1.5\ \mu \rm T$. The measurement was performed here with an output low-pass filter of time constant $\tau=3\ \rm ms$, which gives us a sensitivity $\eta=\sigma \sqrt{\tau}=82\ \rm{nT}/\sqrt{\rm Hz}$.

@@

As a conclusion, we identified three mechanisms causing extra spin depolarization in zero field for dense ensemble of NV$^-$ centers, all related to an increase in the dipole-dipole induced cross-relaxations between the spins of NV centers. The lift in degeneracy of the spin state of different NV classes was found to be the main cause of zero-field depolarization, followed by double-flip processes and then the electric field induced mixing. We have employed cross-relaxation for microwave and orientation-free DC-magnetometry and demonstrated a sensitivity below $100\ \rm{nm}/\sqrt{\rm Hz}$ for a single $10\ \mu \rm m$ crystal and show that that double-flips and electric field play an important role.
Besides magnetometry, our work offers prospects for understanding many body phenomena with strongly coupled spins. 
%DEER, many body, critical thermalization, 2D. Double quantum

>>>>

The electronic spin properties of the negatively charged nitrogen-vacancy (NV$^-$) center in diamond has given rise to a wealth of applications in nanoscale sensing and quantum information science in part thanks to the possibility to optically polarize and read-out its spin state at ambient conditions. Many recent work indeed focus on ensembles of spins from NV centers with the prospect to boost the magnetic field sensing capabilities and for studying many body effects
\citep{TALLAIRE2020421,edmonds2021characterisation, chatzidrosos2021fiberized, kucsko2018critical, giri_coupled_2018}

When the spin ensemble becomes dense, inhomogeneities in the relaxation rates of NVs can give rise to spin depolarisation with a very rich many body dynamics associated with disorder. This mechanisms however limits the efficiency of magnetometers \citep{zhou2020quantum}
%It was proposed to exploit these CR to perform sensitive microwave free magnetometry (REF).  
Recent magnetometry proposals have been put forward at zero magnetic field. There, the depolarisation effect was shown to be much stronger, but the detailed studies of relaxation mechanisms have not been identified to the best of our knowledge. Some of the possible processes that lead to cross-relaxation in strongly coupled dipolar systems are depicted in Fig.~1. The goal of this paper is to identify regimes where the spin flip-flop within different classes of NV centers, double-flip up and down as well as the mixing induced by local electric field play a role.

The electronic spin of NV center is a spin-1 system in the ground state (see Fig. 1). It can be optically polarized in the $\ket{m_s=0}$ state. The photoluminescence of this state is also is larger than the $\ket{m_s=\pm 1}$ states enabling spin-read out at ambient temperature. The $\ket{m_s=\pm 1}$ spin states are separated from the $\ket{m_s=0}$ state by  $D = (2\pi) 2.87$ GHz so that when a resonant microwave or static transverse magnetic field is applied, the photoluminescence is reduced \citep{epstein2005anisotropic,lai2009influence}. 
Fig.  \ref{T1}-a) and b) show the change in photoluminescence (PL) with respect to an external magnetic field for two samples with a low and high concentration of NV$^-$ centers (see SI). Both samples show a decrease in PL as the magnetic field amplitude increases.  There is however a stark difference in the low magnetic field region where only high-density samples shows a drop in PL \citep{jarmola_longitudinal_2015,  mrozek_longitudinal_2015}. This effect can be observed on all our samples (see SI) whenever the NV concentration lies in the ppm range. The drop of the PL at low magnetic field is associated with a decrease of the NV's spin lifetime $T_1$ \citep{jarmola_temperature-_2012}, which results in a decrease of the population in the bright state $\ket{0}$.

%Indeed, since the $\ket{m_s=0}$ spin state is brighter than the $\ket{m_s=\pm 1}$ states, and since the optical pumping in the $\ket{m_s=0}$ state is in competition with the relaxation in a thermal distribution of the states, decreasing the spins lifetime results in a decrease of the NV PL \citep{finco2021imaging}.
%To understand the behavior of the PL in zero-field, we therefore need to study the dynamics of the spin in the low magnetic field region.
The depolarization dynamics of the spin of single or dilute NV centers at room temperature is dominated by two-phonon Raman processes \citep{redman1991spin,jarmola_temperature-_2012,norambuena2018spin}, which depend on the crystal lattice temperature. It has been observed that dense NV centers ensemble have an additional spin decay channel \citep{jarmola_temperature-_2012,jarmola_longitudinal_2015,mrozek_longitudinal_2015, choi2017depolarization, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, pellet2021magnetic, mrozek2021characterization}, which depends greatly on the magnetic field amplitude. This effect has been attributed to cross-relaxation between the NV centers through dipole-dipole coupling \citep{mrozek_longitudinal_2015, choi2017depolarization}.  Inhomogeneity of the spin lifetimes is further needed in order to explain the depolarization of the spin ensemble. We will denote $T_1^{\rm ph}$ the characteristic timescale associated with the phonon relaxation process and $T_1^{\rm dd}$ the density dependent timescale associated with the dipole-dipole relaxation process.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/shema_summary.pdf}
\caption{Schematics showing a diamond with interacting spins as well as the three processes that can account for dipolar relaxation. }
\label{schema_intro}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig_T1.pdf}
\caption{a) Photoluminescence measurement from NV centers ensemble as a function of an external magnetic field applied in an arbitrary direction (a) for the sample CVD-PPB, with [NV$^-$]$\approx 50\ \rm ppb$ (b) for the sample HPHT-150-1, with [NV$^-$]$\approx 3\ \rm ppm$. (c) Sequence used to measure the spin lifetime. (d) Spin relaxation $S_1-S_2$ measured for the dense sample at zero and non-zero magnetic fields. The fitting procedure (orange plain line) is detailed in the main text}
\label{T1}
\end{figure}

@@

Fig. \ref{T1}-c) shows the sequence employed for measuring $T_1^{\rm dd}$ in the dense sample. 
%The spin lifetime protocol used here is described in Fig. \ref{T1} and is based on previous similar experiments 
It consists in a pump-probe measurement where the spins are first polarized in the $\ket{0}$ state by a green laser, and read-out optically after a variable dark time $\tau$. In highly doped samples, this sequence often results in artifacts, mostly due to charge state transfer in the dark \citep{giri_coupled_2018, giri_selective_2019}. It is therefore convenient to repeat the sequence with an additional $\pi$ pulse right before the spin read-out to prepare the remaining $\ket{0}$ polarization into a darker $\ket{+1}$ or $\ket{-1}$ state. By subtracting the result of the two sequences, we select only the spin-dependent part of the signal, with the added benefit of being able to select a specific class of NV centers \citep{jarmola_temperature-_2012, mrozek_longitudinal_2015, choi2017depolarization}.

Fig. \ref{T1}-b) shows the subtracted signals when $\bm B=0$ and $\bm B \neq 0$.
The ensemble spin decay curves are not exponential, which was found to be due to inhomogeneities between the spin lifetimes of dipolar coupled NV$^-$ centers \citep{choi2017depolarization}.
We will base our interpretations of the experimental results on the NV-fluctuator model developed in \citep{choi2017depolarization}. A conclusion of this model is that, for a homogeneous 3D distribution of fluctuators, the dipole-induced lifetime should be stretched-exponential with a stretch factor $\beta=1/2$. Such an ensemble lifetime is indeed observed when $T_1^{\rm dd} \ll T_1^{\rm ph}$.
For the samples used in this paper (detailed in SI) $T_1^{\rm dd} \sim T_1^{\rm ph}$, so both decay processes have to be included in the analysis. Fitting the two curves by the product of an exponential and a stretched exponential, we find $T_1^{\rm ph}=3.6$ ms for both and $T_1^{\rm dd}= 0.6\ \rm ms$ and $13.0\ \rm ms$ respectively. We also conducted further measurements that confirm the lifetime-limited character of the fluctuator employed in the model \citep{choi2017depolarization}, see SI. 
%These results show that our sample clearly manifests dipolar interactions amongst closely packed NV centers, but do not tell about the magnitude of the several possible mechanisms involved in the dipolar relaxation. 
%In spin-1 systems, there can indeed be several channels for decay at zero fields such as flip-flop, double flip,

@@

One of the ways to tune dipole-dipole related phenomena is by changing the number of resonant spins. Fig. \ref{100_VS_1x4} (b)-i) shows an optically detected magnetic resonance (ODMR) spectrum from the dense ensemble HPHT-150-1. The observed lines correspond to the projections of the magnetic fields onto the four "classes" of NV centers - that is physical orientation of the NV axis is the diamond crystal cell - together with the two possible spin transitions $\ket{0}\to\ket{-1}$ and $\ket{0}\to\ket{+1}$. The transitions frequencies of the four different classes can be controlled by changing the amplitude and orientation of the external magnetic field. Fig. \ref{100_VS_1x4} (c)-i) shows an example where all four classes are brought to resonance by placing the magnetic field along the [100] crystalline axis. 
%Other example of class resonances are given in the SI.
This configuration will be essential to decipher what dipolar relaxation mechanisms play a role at low field.

Fig. \ref{100_VS_1x4} (b)-ii) and (b)-iii) show the PL from an ensemble of NV centers as a function of the amplitude of a magnetic field aligned as in Fig. \ref{100_VS_1x4} (b)-i). 
We also extract the stretched exponential spin decay time $T_1^{\rm dd}$ by running the sequence shown in Fig. 2-c) at each of the 0.5 G magnetic field values. 
We observe that the 3 \% increase in PL under low magnetic field is indeed correlated with a drop in the spin decay rate from 1.6 kHz to 0.08 kHz, as the magnetic field increases. This result can be explained by the higher number of resonant spins in zero magnetic field, where all four classes are degenerate, as opposed when $\bm B \neq 0$. The half width of the dip ($\approx 10$G) is consistent with a fluctuator model taking into account the change in the spin resonance overlaps as the $B$ field is swept (see SI ?) [Je sais pas, je le fais ?].
Note that the decrease in PL for $B>40\ \rm G$ are related to state mixing by the transverse magnetic field, so it is not correlated to a modification of the spin lifetime.

Importantly, we observed that the lifted degeneracy of the four NV classes is not the sole reason for the increase of the spin lifetime as the B field increases. Fig. \ref{100_VS_1x4} (c)-ii) and (c)-iii) show the change in the PL and in $1/T_1^{\rm dd}$ as a function of a magnetic field that is aligned along the [100] axis. For this particular orientation, the four classes are always resonant, regardless of the field amplitude. We can see that there still is an increase in the spin decay rate and a corresponding drop in the PL under low magnetic field values, although considerably reduced. 
%Also, the rate of change in both the PL and $T_1^{\rm dd}$ is larger by more than a factor of 3 in the [100] configuration. Although the contrast is smaller, this effect can thus drastically impact zero-field magnetometry protocols based on CR. C'est faux ça, c'est meme plutot le contraire ici
The main aim of this paper is to understand the physics governing the contribution to the spin decay in this second scenario. Note that the slight drop in PL and the corresponding bump for $1/T_1^{\rm dd}$ at $B \sim 20$G is related to dipolar interaction with NV centers that have a $^{13}$C as a first neighbor \cite{pellet2021optical}.

We identified and isolated two additional mechanisms that could explain this effect : local electric fields and double-flip processes. 
The dipole-dipole interaction Hamiltonian between two spins ${\vec S}_1$ and ${\vec S}_2$ reads :
\begin{equation}
\mathcal{H}^{\rm dd}= -\frac{J_0}{r^3}\left(3\left({\vec S}_1 \cdot \vec u \right)\left({\vec S}_2 \cdot \vec u \right) - {\vec S}_1 \cdot {\vec S}_2  \right),
\end{equation}
Where $J_0= (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\vec r$ is the relative positions of the two spins and $\vec u = \frac{\vec r}{\norm{\vec r}}$. In situations such as the one described in Fig.XX, the only relevant terms of $\mathcal{H}^{\rm dd}$ in terms of population transfer are the flip-flop terms such as $\mel{0,+1}{\mathcal{H}^{\rm dd}}{+1,0}$. In zero magnetic field however, we have to take into account these two other mechanisms of the dipolar Hamiltonian in consideration.
%where both NV spins involved in the dipole-dipole coupling flip their spin in the same direction.
It was shown in  \cite{mittiga2018imaging} that local electric fields coming from charge traps such as $P_1^+$ and $NV^-$ centers are responsible for the optically detected spin resonance profile at zero magnetic field in large NV ensembles. Magnetic field noise comes in to second order in zero field. %We also do not take into account the hyper-fine structure of the NV center due to the large inhomogeneous broadening of the ESR transitions (see SI). 
We will then consider the following spin Hamiltonian for the electric field dependent NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_{\rm elec}= d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
where $E_{x,y}$ are the projections of local electric fields on the NV axes. 
Under local electric fields with orientations given by the angle $\phi_E={\rm tan}(E_x/E_y)$, the eigenstates of $H_{NV}$ are $\ket{0}$ and $\ket{\pm}=\frac{1}{\sqrt{2}}(\ket{+1}\pm e^{-i\phi_E}\ket{-1})$.
Computing the flip-flop terms $\ket{\pm,0} \bra{0,\pm} $ in the fluctuator model shows that $1/T_1^{dd}$ should be increased at zero field, even after averaging over all possible orientations $\phi_E$ (see SI).  Another mechanism at stake when considering dipolar interactions with spin-1 systems are double-flip processes. 
These processes correspond to terms like $\ket{+1,0} \bra{0,-1}$ in the anisotropic (?) dipolar interaction, giving rise to an exchange of two units of spin-angular momentum. These processes can become significant when the states $\pm 1$ are degenerate and are thus likely to explain part of the zero-field dip in Fig. c)-ii). 
%Theoretical model predicts a factor of $\approx $5 increase in $1/T_1^{dd}$ at zero field.

@@

Crucially, mixing remains even when the associated energy splitting of the $\ket{d}$ and $\ket{e}$ states is large : Fig. \ref{B_transverse}-(b) shows both the splitting between the energy levels of $\ket{d}$ and $\ket{e}$, and the matching factor $\abs{\bra{e}\ket{+}}^2$ characterizing the closeness of the states $\bra{e}$ and $\bra{+}$. We can see that at a field value of $\sim 130$ G, the splitting between the $\ket{d}$ and $\ket{e}$ reaches $\sim 50$ MHz, far exceeding the range of the dipole-dipole resonant coupling (see details in SI), but $\abs{\bra{e}\ket{+}}^2 > 0.98$ meaning that the spin Hamiltonian eigenstates are still roughly in the $\ket{0}, \ket{+}, \ket{-}$ basis. We can therefore decouple the effects of the double-flip processes  $\propto \ket{g,d}\bra{d,e}$, which are no longer resonant when $| \bm B | \geq 100$ G, from the effects of the mixing induced by the transverse electric - or in this case magnetic - field.

Fig. \ref{B_transverse}-(c) shows the measurement of the stretched lifetime $T_1^{\rm dd}$ for a single class of NV centers exposed to a transverse magnetic field between 20 and 130 G on another sample HPHT-150-2, upon which we measured a stretched decay $1/T_1^{\rm dd}=25\pm 5\  \rm Hz$ in the case of strong longitudinal magnetic field. The detuning $\Delta \nu$ between the two states $\ket{d}$ and $\ket{e}$ was measured through ODMR. We attribute the decrease in the spin decay rate when the detuning increases to the loss of effectiveness of the double-flip processes. We can notice that the decay rate reaches a plateau for $\Delta \nu \geq 30\ \rm MHz$ with a value $1/T_1^{\rm dd}=57\pm 5\  \rm Hz$ about twice as large as the decay rate for a longitudinal magnetic field. This plateau we attribute to the effect of the transverse magnetic field, which we assume to be similar to the effect of the electric field in the low magnetic field regime. We should note that the effect of the double-flip processes is about 5 times more important in the depolarization rate than the effect of the electric field, even though the states are not initially fully resonant.

@@

Fig. \ref{magneto} (a) shows the demodulated PL as a DC magnetic field is scanned in a random direction and an additional alternating magnetic field of amplitude $\sim 10\ \rm G$ and frequency  $\sim 1\ \rm kHz$ is added through the same electromagnet. The sample used here is HPHT-15-1 and the laser power $\sim 1\ \rm mW$.  We can see a sharp linear slope in low field $\abs{B} < 5\ \rm G$. Once calibrated, in this case with ODMR, the slope can provide a 1D magnetic field measurement, which could be extended to 3D with a set of 3 coils or 3 electromagnets, as in \cite{zheng_microwave-free_2020}. In order to assess the sensitivity of the measurement, we alternate a small DC field of $\approx 40\ \mu\rm{T}$ every few seconds and take a histogram of the measured fields, as shown in Fig. \ref{magneto} (b) and (c). The histogram is well fitted with gaussians of standard deviation $\sigma=1.5\ \mu \rm T$. The measurement was performed here with an output low-pass filter of time constant $\tau=3\ \rm ms$, which gives us a sensitivity $\eta=\sigma \sqrt{\tau}=82\ \rm{nT}/\sqrt{\rm Hz}$.

@@

As a conclusion, we identified three mechanisms causing extra spin depolarization in zero field for dense ensemble of NV$^-$ centers, all related to an increase in the dipole-dipole induced cross-relaxations between the spins of NV centers. The lift in degeneracy of the spin state of different NV classes was found to be the main cause of zero-field depolarization, followed by double-flip processes and then the electric field induced mixing. We have employed cross-relaxation for microwave and orientation-free DC-magnetometry and demonstrated a sensitivity below $100\ \rm{nm}/\sqrt{\rm Hz}$ for a single $10\ \mu \rm m$ commercially available diamond and show that that double-flips and electric field play an important role.
Besides magnetometry, our work offers prospects for understanding many body phenomena with strongly coupled spins. 
%DEER, many body, critical thermalization, 2D. Double quantum
